PanGraph: scalable bacterial pan-genome graph construction

### 2023-02-23 reviewer_response.tex

      Noll et al set out to build a representation which combines a global pan-genome with local MSAs, and to do so in a way which scales well, and which provides easy to use visualisations. It is an ambitious aim, and in my opinion this is great work. They have developed something noone else has (see enumeration above), and the paper does a good job of describing and then evaluating their work - they are to be congratulated. This tool is sure to be extremely useful, and very valuable to the community. I have no serious reservations about the quality or calibre of the work, i admire it greatly.\\

      Nevertheless, there are a number of clarifications which need to be addressed before the paper is suitable for publication.}

\subsubsection*{Major Issues}

\reviewer{R2.1}{Line 46: when you say "PanGraph transforms an arbitrary set of genomes into a graph that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms." - I am a bit unclear on how the nucleotide level variation is summarised. The paper is a bit hand-wavey about this. When i dug around on github , i found this helpful page \url{https://neherlab.github.io/pangraph/tutorials/tutorial_2/} Which answers and raises questions. So, my question: for each pancontig (which i think is the same as a block), i think what you do is infer a consensus (what is this? majority base at each position but somehow also choose an allele when there is an indel?), and then you describe the other alleles in the MSA as edits from this? You surely do not do all pairwise descriptions of SNPS, right? On that same webpage you have a section on ``block consensus to node alignment'' - how do you choose the order of alleles here? This is all valuable information and it would be great to have it in the supplementary or main methods somewhere.}
% \Marco{this point was also raised by R3. Mutations are always represented as variation on the consensus. The way consensus is propagated during graph merging might require more explanation. Maybe we could add a section in the supplementary with a somewhat more detailed description of the algorithmic steps we perform when merging two graphs? In the code \href{https://github.com/neherlab/pangraph/blob/e1c11012a3d19975d4b1dc61b1af3eb3949f7cd2/src/block.jl\#L1274}{this part} takes care of merging alignments between two blocks with only small (i.e. below cutoff) insertions and deletions. In short what we do is stitch the two alignments together using the cigar string as a guide, and re-compute the consensus on the newly-obtained alignment. This ensures that the consensus agrees with the majority of sequences in the alignment, but does not solve the full multiple sequence alignment problem. However we also provide the ``polish'' step for this, that can be executed after graph creation. This will take care of refining the block alignments.}\\
% \Liam{I think a `narrative section' in the supplementary describing this in brief sounds great.}
\response{In our pangenome graph representation each pancontig encodes an alignment using a consensus sequence, and encoding each line as variations over the consensus. The consensus sequence is defined using the majority rule on the alignment and is updated at each pancontig merge. Each line in the alignment is then only described as variations over this consensus (mutations, insertions and deletions).\\
      To make this more clear we introduced section I in SI, in which we explain in more detail the merging procedure and how the consensus sequence is updated. We also refer to the documentation page mentioned by the reviewer. Moreover, to avoid ambiguity we also avoided referring to pancontigs as ``blocks'' in the main text.}

\reviewer{R2.2}{Line 61: I agree and understand how each sequence can be sketched in linear time. I don't follow how you calculate all pairwise intersections through sorting the list of minimizers, nor do i see how this ends up being log-linear. For each genome, you can sort it's minimizers in nlog(n) (ie log linear), where n ~genome length. How do you get all pairwise distances from that without another multiplier of n?}
% \Marco{in the text we write that ``the pairwise distance matrix is estimated in a time that increases log-linearly in the \textbf{total length} of sequence that is analyzed.'' This might generate some confusion since the scaling depends both on the number of genomes $N$ and their length $l$. The algorithmic time for all the pairwise distances should scale as $\sim N^2 l \log l$. We should be in a regime where $N$ is not a limiting factor, it's just a multiplicative constant, and thus I think our statement is fundamentally correct. But we can make it more precise. This scaling is at any rate not a limiting factor in pangraph.}\\
% \Liam{clarifying $N$ and $l$ makes sense.}
\responsedraft{We agree with the reviewer that the formulation in the main text is too concise and not completely correct. There are two main variables at play: the length of a genomes $l$ and the number of isolates $n$. As the reviewer points out, the set of minimizers for each genome can be created and sorted in $l \log l$. Finding the set of shared minimizers would require comparing $\sim n^2$ sets of ordered lists of minimizers, which scales as $\sim n^2 l$. This would provide a scaling $O(n l \log l + n^2 l)$.\\
      In our implementation we achieve a slightly better scaling as a function of $n$ by sketching and sorting all the genomes together, which can be done in time $L \log L$, where $L= n \times l$ is the total length of all genomes. We keep the information on each minimizer origin in the list. We then scroll the list and for each chunk of equal minimizers we update the corresponding entries of a $n^2$ matrix of minimizer hit counts. Depending on the amount of shared minimizers, this scales as $n^2 l$ in the worst case (all minimizers shared) and as $n l$ in the best case (no minimizer shared). Therefore depending on the diversity of the sample the scaling will be between $O(L \log L + n L)$ and $O(L \log L)$.\\
      In practice for typical use cases of pangraph ($n \sim 100$ and $L \sim 10^6$) the computationally challenging part is the minimizer sketching, and not the distance matrix computation. This is partly due to some prefactors (e.g. the window size of 100 bp used for sketching makes so that the list of minimizer is significantly shorter than the genome, and the only operation performed $n^2$ times is a single increment of an integer in the minimizer hit matrix). We also stress that the computational time to build the guide tree is usually negligible compared to the time required for the graph merging procedure.\\
      To correct this ambiguity we...
      % missing: changes in main text. My plan: report the n^2 l log l scaling. 
}

\reviewer{R2.3}{The section on iterative graph alignment (lines 69-77) is in one sense quite clear and intuitive, but there is a lot of detail missing which leaves the reader a bit confused when they try to think about details. I would appreciate some detail being added, perhaps in the supplementary methods. For example:
      \begin{itemize}
            \item at the start you just have two genomes, and you scan them for orthologous regions, and you decide which to merge. As i understand it, you merge them in order of pseudoenergy ranking, which makes sense (what do you do for precise duplicates, where a region on genome 1 is a precise match for 2 regions on genome 2? random choice?)
                  % \Marco{the ranking would be the same and we perform only one of the two merges in the first phase, when comparing the two graphs. After this we align the graph on itself and the second merge should be performed at this point. This comment also suggests we could add a section in the supplementary with a more detailed description of the algorithm?}
            \item as you move on, you are aligning graphs, and you really skim over the issue of how you do that. How does minimap or mmseqs allow you to look for homology between two graphs? The only thing i can imagine you are doing is only looking for homology between the genomes in one graph, and the genomes in the second graph. I would really appreciate it if you could clarify this and even ideally add some example figures in the supplement showing with some toy genomes how you align two graphs.
                  % \Marco{another comment pointing at the fact that we could add a more detailed explanation about graph merging on the supplementary. What we do is using the aligner using as reference and queries the consensus sequences of each block in the two different graphs, and create a first version of the merged graph. We then repeat this procedure in an all-vs-all comparison of the consensus sequences of the blocks, and proceed with further merges. This is necessary because when two alignments overlap only one of the two is performed in the first alignment round, but the second will be performed in further rounds.}\\
                  % \Liam{agree, sounds like a section in supplementary is a good idea to address these comments.}
      \end{itemize}}
\response{We followed the advice of the reviewer and added Section I in the SI, in which we expand on the explanation of how we merge two graphs.\\
      In short, we perform the merging procedure in two steps. In the first we use the aligner (minimap2 or mmseqs2) to scan for homology matches between the consensus sequence of the pancontigs of one graph and of the other. These matches are ranked by pseudo-energy and performed in order. Each pancontig can be subject to only one merge at this stage.\\
      To answer the first point, for precise duplicates only one of the two matches is merged at this stage, with pseudo-random ordering. However also the second region will be matched and merged before the merging of the two graphs is conlcuded.\\
      After all of the matches have been exhausted, the two graphs have undergone a preliminary merging, with some new merged pancontigs being created. At this point we repeat the procedure looking for matches on the consensus sequence of all the pancontigs in an all-against-all comparison. This self-mapping on the graph will detect any homology that could not have been merged at the previous stage, and will perform the corresponding merging. The self-mapping is repeated iteratively until no new merges can be performed.
}

\reviewer{R2.4}{Line 92-94, same section. You give two statements:
      ``Only mergers whose alignment has negative pseudo-energy are performed''
      and
      ``no mergers will be performed if the Hamming distance exceeds $1/\beta$.''
      These are separate criteria that prevent mergers, right? So any potential merger with negative pseudoenergy that does not fail these criteria will happen, with the caveat that you perform mergers in order?\\
      One interesting thing here is that you are effectively merging things in reverse evolutionary order, as you have lower energy for longer more similar things which presumably diverged more recently. Not sure if that tells us anything.}
% \Marco{the second is actually a consequence of the first, but except for that it is correct.}\\
% \Liam{maybe say `$\beta$ controls the maximal sequence divergence between the two merger-candidates. The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.'}\\
% \Marco{indeed, the parameter $\beta$ could control how far in the past one wants to accept merges. Should we mention this explicilty in the text?}\\
% \Liam{could be mentioned as an `intuitively, this should correspond to merging of homologous stretches of sequence in roughly reverse evolutionary order'.}\\
\response{
      In this case the second criterion is a consequence of the first. From the definition of the pseudo-energy (cf. eq.~(1) in the main text) it follows that for an alignment of pancontig consensus sequences to have negative pseudo-energy must be that the number of SNPs divided by the alignment length $N_m/l$ must be less than $1/\beta$. Moreover, as the reviewer pointed out, the pseudo-energy ranking of merges effectively results in merges being performed in order of similarity, effectively in reverse evolutionary order.\\
      To make this more explicit we modified lines 93-95 in ``$\beta$ controls the maximal sequence divergence between the two merger-candidates. The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$... The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.''
      % QUESTION: is there ambiguity in the main text on the value of N_m? As SNPs in the alignment? Misleading figure?
}

@@

\reviewer{R2.5}{I very much appreciated that you measure the accuracy of pangraph via this displacement idea. I started writing a kindly worded criticism that I thought your method was flawed - surely the displacement depends not only on the order of doing merges, and decisions made about what to do when there are two equally good places to merge with. I guess i was saying that slightly different evolutionary events could give the same genomes, and thereby justify the errors. However i very much like the approach you detail in the methods, numerically solving the assignment problem. So, this started as a major issue, but in the end is a compliment.}

\reviewer{R2.6}{Line 183 - I really do not understand what is going on with your alpha,beta being set to zero. Are you saying if you stop penalising merges for having too many SNPs, then you will be able to merge more, and this will stop you from havign a 10\% divergence limit? It will also mean you make more false merges, right?}
% \Marco{setting $\beta$ to zero will indeed remove the 10\% divergence limit set by the pseudo-energy. There is another limit that depends on the aligner sensitivity (20 to 30\%, depending on the aligner). This limit remains, and in this sense no ``false merges'' are performed? Not sure I interpret this correctly.}\\
% \Liam{I think your interpretation is right. Important to clarify that there are those two limits and that $\beta$ only controls the divergence limit \textit{within} what one can actually align.}
\response{There are two factors limiting the divergence of merges that are performed. The first factor is the aligner used. Minimap2 is in general able to find matches up to roughly 20\% sequence divergence, while mmseqs2 can reach divergence up to roughly 30\%. Only homology matches of up to this divergence will be suggested by the aligners. On top of this, only matches with negative pseudo-energy are turned into merges. For any given value of $\beta$, this sets an additional limit to the divergence, which must be less than $1/\beta$.\\
      This is now made more explicit in SI section I (PanGraph parameters effect).\\
      When setting $\beta = 0$ only the intrinsic limit on divergence posed by the aligner sensitivity is left. In this sense more merges are allowed, but all mergess will have some level of homology (maximum 20\% or 30\% divergence, depending on the aligner).}

\reviewer{R2.7}{Why does the L50 peak go up for null-scores for mmseqs for Mtb, Pm, Kpn and Ec (Figure 4E, purple with lines) but down for helicobacter?}
% \Marco{the effect of merging more blocks on L50 is not trivial. It depends on whether the merging occurs between long or short blocks, and how much fragmentation this causes. HP is a special case in that its diverge is on the threshold of what minimap2 can handle. Using mmseqs2 results in a pangenome that is half the size compared to standard parameters (fig. 4G). This is an impressive amount of additional compression. At the same time the fraction of core pangenome more than doubles (4H). The total number of blocks decreases to roughly half (4D) while N50 increases (4F). For me this points to the fact that in the case of HP merging will split long blocks and create more fragmentation. For the reviewer I could produce the block frequency and lenght distribution for the two cases and comparing them would give an answer. I could do the same for EC or KP for comparison.}\\
% \Liam{makes sense.}
\response{L50 measures the minimum number of pancontigs needed to cover at least 50\% of the pangenome graph sequence. This number is closely related to the total number of pancontigs. In general the use of null energy parameters result in more fragmented graphs, with a higher number of pancontigs (cf. fig.~4D in SI). Hp is an exception and has the opposite trend, both for the number of pancontig and for the value of L50. For this species the use of null energy parameters results in less fragmentation of the graph. This is compatible with the observation that the average divergence of this dataset is on the limit of what PanGraph can handle. Unless a highly sensitive aligner is used and the divergence barrier in pseudo-energy is removed, some highly-diverged homologous sequences remain split in different pancontigs. Conversely, the choice of these parameters generates a graph with less pancontigs, a lower L50, a larger fraction of core pangenome, an higher median length of pancontigs (N50).
      % TODO: add mention in SI
      % TODO: plot change for HP and KP/EC as comparison?
      % QUESTION: this is more or less explicit in SI? Should I expand more?
}

\reviewer{R2.8}{Just to make sure i understand - is it possible for sequence to be ``left out'' in between pancontigs?}
\Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}\\
\Liam{if they possibly misunderstood, suggests that some statement that all sequence ends up in a pancontig should be added to manuscript. I suspect they are thinking of case of short ($<$100bp) stretches `in between' pancontigs, which (if I understand correctly) will be incorporated as insertions at the end of the consensus alignment of the adjacent pancontig.}

### 2023-02-23 main.tex
chore: update

\documentclass[aps,rmp,preprint,superscriptaddress,10pt,linenumbers]{revtex4-1}
%\documentclass[aps,rmp,reprint,superscriptaddress,10pt]{revtex4-1}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsthm,amsfonts,amssymb,amscd}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{enumerate}
\usepackage{color}
\usepackage[final]{hyperref}

\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\Lcore}{L_\text{core}}
\newcommand{\Lsoftcore}{L_\text{95\%}}
\newcommand{\Lpang}{L_\text{pang}}
\newcommand{\Lgen}{L_\text{gen}}
\newcommand{\dcore}{\langle d_\text{core} \rangle}
\newcommand{\Richard}[1]{{\color{red}Richard: #1}}

@@

\paragraph*{Merging of homologous sequence.}
If the above alignment step detected homologous stretches between two pancontigs, these pancontigs, or parts of them, can be merged.
It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.
We rank each alignment between two pancontigs according to the pseudo-energy
\begin{equation}\label{eq:pseudo-energy}
    E = -\ell + \alpha N_c + \beta N_m
\end{equation}
where $\ell$, $N_c$, and $N_m$ denote the alignment length, number of \emph{pancontigs} created by the merger, and number of polymorphisms per genome in the newly created {pancontig} respectively.
See Fig \ref{fig:visualization}B for a graphical definition for each term.
Additionally, $\alpha$ and $\beta$ are hyperparameters of the algorithm.
Only mergers whose alignment has negative pseudo-energy are performed.
The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates and no mergers will be performed if the Hamming distance exceeds $1/\beta$.
Importantly, this divergence threshold is applied to the comparison of the consensus sequences of the merger candidates.
The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the resulting pan-contigs resulting from the merger.
The merger is always performed if the merger doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.

@@

The above procedure is repeated until no alignments with negative energy remain.
Upon completion, transitive edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent \emph{pancontigs}.

@@

\begin{figure}[tb]
    \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
    \caption{{\bf Accuracy against synthetic data}.
        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels \textit{minimap2} with \textit{asm10} or \textit{asm20} option, and \textit{mmseqs2}. In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp. The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of 100bp, as a function of average pairwise sequence divergence. Line and shaded area represents mean and standard deviation over 25 repetitions. \textit{mmseqs2} maintains accuracy at higher divergence, at the cost of higher computational time.
    }
    \label{fig:toy-accuracy}
\end{figure}

@@

\begin{table}[h]
    \setlength{\tabcolsep}{6pt}
    \begin{tabular}{l c c c c c c}
        \hline\hline
        Species                  & N.  & $\avg{\Lgen}$ & $\Lpang$ & $\Lcore$ & $\Lsoftcore$ & $\dcore$ \\
                                 &     & [Mbp]         & [Mbp]    & [Mbp]    & [Mbp]        & \\
        \hline
        \textit{E.~coli}         & 307 & 5.0           & 17.7     & 0.7      & 2.0          & 1.6\%    \\
        \textit{K.~pneumoniae}   & 109 & 5.3           & 13.0     & 2.3      & 3.7          & 0.8\%    \\
        \textit{H.~pylori}       & 85  & 1.6           & 1.9      & 0.6      & 1.0          & 4.2\%    \\
        \textit{M.~tuberculosis} & 51  & 4.4           & 4.1      & 2.4      & 3.4          & 0.03\%   \\
        \textit{P.~marinus}      & 10  & 1.8           & 3.3      & 0.7      & 0.7          & 26.9\%   \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset properties}. Columns represent:
        number of isolates (N.),
        average chromosome length ($\avg{\Lgen}$),
        total pangenome length ($\Lpang$),
        total core genome length ($\Lcore$),
        total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates), and
        average pairwise divergence of core genes ($\dcore$).
    }
    \label{table:panx-dataset}
\end{table}

@@

In addition to being a useful operation, marginalization allows us to quantify the consistency and robustness of pangraph construction on real data, where the ground truth is unknown.
To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{K.~pneumoniae} dataset. We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair (i.e. marginalized graph in Fig.~\ref{fig:marginalization}A top), and the graph built directly from the pair (i.e. pairwise graph in Fig.~\ref{fig:marginalization}A top).
In order to compare both graphs, we compute the partition they generated on the genome of the isolates they include. Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate (cf. Fig.~\ref{fig:marginalization}A bottom). We can classify the segments in the intersection of the two partitions in different categories, depending on whether the two partitions agree on whether the segment is shared or private. Moreover segments on which the two partitions agree are further sub-divided depending on whether they are private or shared on both partitions.\\
The compatibility between the two graphs requires segments on which the two partitions disagree to be few and short. Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} (cf. Fig.~\ref{fig:marginalization}B-C). Conversely, segments on which the partitions agree have average size of several kbps. The fact that the two partitions almost completely agree cannot be explained simply by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence. We confirmed these results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}. Namely, we approximate the fraction of homologous sequence between any two pairs using the fraction of shared k-mers ($k=21$) divided by $(1-d)^{k}$, where $d$ is the average pairwise divergence on core genes for the pair considered. This divisor corrects for k-mers on homologous sequence that are not shared due to mutations. The resulting distribution of shared sequence is in very good agreement with the fraction of segments that are shared on both graphs (cf. Fig.~\ref{fig:marginalization}B), suggesting that homologous sequences are correctly merged on the complete, marginalized and pairwise graphs.\\
We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf. SI Section IV and SI Fig.~5).

>>>>

\documentclass[aps,rmp,preprint,superscriptaddress,10pt,linenumbers]{revtex4-1}
%\documentclass[aps,rmp,reprint,superscriptaddress,10pt]{revtex4-1}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsthm,amsfonts,amssymb,amscd}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{enumerate}
\usepackage{xcolor}
\usepackage[normalem]{ulem}
\usepackage[final]{hyperref}

\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\Lcore}{L_\text{core}}
\newcommand{\Lsoftcore}{L_\text{95\%}}
\newcommand{\Lpang}{L_\text{pang}}
\newcommand{\Lgen}{L_\text{gen}}
\newcommand{\Lthr}{L_{\min}}
\newcommand{\dcore}{\langle d_\text{core} \rangle}
\newcommand{\modified}[1]{{\color{red}#1}}
\newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}

@@

\paragraph*{Merging of homologous sequence.}
If the above alignment step detected homologous stretches between \oldnew{}{the consensus sequences of} two pancontigs, these pancontigs, or parts of them, can be merged.
It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.
We rank each alignment between two pancontigs according to the pseudo-energy
\begin{equation}\label{eq:pseudo-energy}
    E = -\ell + \alpha N_c + \beta N_m
\end{equation}
where $\ell$, $N_c$, and $N_m$ denote the alignment length, number of \emph{pancontigs} created by the merger, and number of polymorphisms per genome in the newly created {pancontig} respectively.
See Fig \ref{fig:visualization}B for a graphical definition for each term.
Additionally, $\alpha$ and $\beta$ are hyperparameters of the algorithm.
Only mergers whose alignment has negative pseudo-energy are performed.
The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates\oldnew{ and no mergers will be performed if the Hamming distance exceeds $1/\beta$.}{. The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.}
Importantly, this divergence threshold is applied to the comparison of the consensus sequences of the merger candidates.
\oldnew{}{The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.}
The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the \oldnew{resulting}{} pan-contigs resulting from the merger.
The merger is always performed if \oldnew{the merger}{it} doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
\oldnew{In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.}{In addition to the fragmentation parameter $\alpha$, there is a further parameter $\Lthr$ that controls the minimal length of pancontigs. Structural variation at a scale smaller than this threshold (i.e. short indels) is stored in pancontig alignments, rather than in the graph structure, and no merges are performed if they have length shorter than this threshold. We set a default value of 100 bp for this parameter.}

@@

The above procedure is repeated until no alignments with negative energy remain.
Upon completion, transitive edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent \emph{pancontigs}. \oldnew{}{More details on the merging procedure and the effect of parameters can be found in SI section I.}

@@

\begin{figure}[tb]
    \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
    \caption{{\bf Accuracy against synthetic data}.
        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels \textit{minimap2} with \textit{asm10} or \textit{asm20} option, and \textit{mmseqs2}. In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp. The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of \oldnew{}{$\Lthr=$}100bp, as a function of average pairwise sequence divergence. Line and shaded area represents mean and standard deviation over 25 repetitions. \textit{mmseqs2} maintains accuracy at higher divergence, at the cost of higher computational time.
    }
    \label{fig:toy-accuracy}
\end{figure}

@@

\begin{table}[h]
    \setlength{\tabcolsep}{6pt}
    \begin{tabular}{l c c c c c c}
        \hline\hline
        Species                  & N.  & $\avg{\Lgen}$ & $\Lpang$ & $\Lcore$ & $\Lsoftcore$ & $\dcore$ \\
                                 &     & [Mbp]         & [Mbp]    & [Mbp]    & [Mbp]        &          \\
        \hline
        \textit{E.~coli}         & 307 & 5.0           & 17.7     & 0.7      & 2.0          & 1.6\%    \\
        \textit{K.~pneumoniae}   & 109 & 5.3           & 13.0     & 2.3      & 3.7          & 0.8\%    \\
        \textit{H.~pylori}       & 85  & 1.6           & 1.9      & 0.6      & 1.0          & 4.2\%    \\
        \textit{M.~tuberculosis} & 51  & 4.4           & 4.1      & 2.4      & 3.4          & 0.03\%   \\
        \textit{P.~marinus}      & 10  & 1.8           & 3.3      & 0.7      & 0.7          & 26.9\%   \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset properties}. Columns represent:
        number of isolates (N.),
        average chromosome length ($\avg{\Lgen}$),
        total pangenome length ($\Lpang$),
        total core genome length ($\Lcore$),
        \oldnew{}{total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates),}
        and average pairwise divergence of core genes ($\dcore$).
    }
    \label{table:panx-dataset}
\end{table}

@@

In addition to being a useful operation, marginalization allows us to quantify the consistency and robustness of pangraph construction on real data, where the ground truth is unknown.
To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{K.~pneumoniae} dataset. We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair (i.e. marginalized graph in Fig.~\ref{fig:marginalization}A top), and the graph built directly from the pair (i.e. pairwise graph in Fig.~\ref{fig:marginalization}A top).
In order to compare both graphs, we compute the partition they generated on the genome of the isolates they include. Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate (cf. Fig.~\ref{fig:marginalization}A bottom). We can classify the segments in the intersection of the two partitions in different categories, depending on whether the two partitions agree on whether the segment is shared or private. Moreover segments on which the two partitions agree are further sub-divided depending on whether they are private or shared on both partitions.\\
The compatibility between the two graphs requires segments on which the two partitions disagree to be few and short. Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} \oldnew{}{set by the value of $\Lthr$} (cf. Fig.~\ref{fig:marginalization}B-C). Conversely, segments on which the partitions agree have average size of several kbps. The fact that the two partitions almost completely agree cannot be explained simply by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence. We confirmed these results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}. Namely, we approximate the fraction of homologous sequence between any two pairs using the fraction of shared k-mers ($k=21$) divided by $(1-d)^{k}$, where $d$ is the average pairwise divergence on core genes for the pair considered. This divisor corrects for k-mers on homologous sequence that are not shared due to mutations. The resulting distribution of shared sequence is in very good agreement with the fraction of segments that are shared on both graphs (cf. Fig.~\ref{fig:marginalization}B), suggesting that homologous sequences are correctly merged on the complete, marginalized and pairwise graphs.\\
We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf. SI Section IV and SI Fig.~5).
